---
title: "Lesson 12: Multicollinearity & Other Regression Pitfalls"
output: html_notebook
---
https://onlinecourses.science.psu.edu/stat501/node/430/

```{r}
library(ggplot2)
```

# Blood pressure (multicollinearity)

* Load the bloodpress data.
* Create scatterplot matrices of the data.
* Calculate correlations between the variables.
```{r}
bloodpress <- read.table("DataL12/bloodpress.txt", header=T)

pairs(bloodpress[,c(2:5)])
```
```{r}
pairs(bloodpress[,c(2,6:8)])
```
```{r}
round(cor(bloodpress[,c(2:8)]),3)
```
# Uncorrelated predictors (no multicollinearity)

* Load the uncorrpreds data.
* Create a scatterplot matrix of the data.
* Calculate the correlation between the predictors.
* Fit a simple linear regression model of y vs x1.
* Fit a simple linear regression model of y vs x2.
* Fit a multiple linear regression model of y vs x1 + x2.
* Fit a multiple linear regression model of y vs x2 + x1.
* Use the scatter3d function in the car package to create a 3D scatterplot of the data with the fitted plane for a multiple linear regression model of y vs x1 + x2.

```{r}
uncorrpreds <- read.table("DataL12/uncorrpreds.csv", sep=",", header=T)
pairs(uncorrpreds)
```
```{r}
with(uncorrpreds, cor(x1,x2))
```
```{r}
model.1 <- lm(y ~ x1, data=uncorrpreds)
summary(model.1)
```
```{r}
anova(model.1)
```
```{r}
model.2 <- lm(y ~ x2, data=uncorrpreds)
summary(model.2)
```
```{r}
anova(model.2)
```
```{r}
model.21 <- lm(y ~ x2 + x1, data=uncorrpreds)
summary(model.21)
```
```{r}
anova(model.21)
```
```{r message=FALSE}
library(car)
head(uncorrpreds)
attach(uncorrpreds)
scatter3d(y ~ x1 + x2)
detach(uncorrpreds)
```
# Blood pressure (predictors with almost no multicollinearity)

* Load the bloodpress data.
* Create a scatterplot matrix of the data.
* Fit a simple linear regression model of BP vs Stress.
* Fit a simple linear regression model of BP vs BSA.
* Fit a multiple linear regression model of BP vs Stress + BSA.
* Fit a multiple linear regression model of BP vs BSA + Stress.
* Use the scatter3d function in the car package to create a 3D scatterplot of the data with the fitted plane for a multiple linear regression model of BP vs Stress + BSA.
```{r}
bloodpress <- read.table("DataL12/bloodpress.txt", header=T)
pairs(bloodpress[,c(2,5,8)])
```
```{r}
model.1 <- lm(BP ~ Stress, data=bloodpress)
summary(model.1)
anova(model.1)
```
```{r}
model.2 <- lm(BP ~ BSA, data = bloodpress)
summary(model.2)
anova(model.12)
```
```{r}
model.21 <- lm(BP ~ BSA + Stress, data = bloodpress)
summary(model.21)
anova(model.21)
```
```{r}
library(car)
attach(bloodpress)
scatter3d(BP ~ Stress + BSA)
detach(bloodpress)
```
# Blood pressure (predictors with high multicollinearity)

    Load the bloodpress data.
    Create a scatterplot matrix of the data.
    Fit a simple linear regression model of BP vs Weight.
    Fit a simple linear regression model of BP vs BSA.
    Fit a multiple linear regression model of BP vs Weight + BSA.
    Fit a multiple linear regression model of BP vs BSA + Weight.
    Use the scatter3d function in the car package to create a 3D scatterplot of the data with the fitted plane for a multiple linear regression model of BP vs Weight + BSA.
    Predict BP for Weight=92 and BSA=2 for the two simple linear regression models and the multiple linear regression model.
```{r}
pairs(bloodpress[,c(2,5,4)])

```
```{r}
model.1 <- lm(BP ~ Weight, data = bloodpress)
summary(model.1)
anova(model.1)
```
```{r}
model.2 <- lm(BP ~ BSA, data = bloodpress)
summary(model.2)
anova(model.2)
```
```{r}
model.12 <- lm(BP ~ Weight + BSA, data=bloodpress)
summary(model.12)
anova(model.12)
```
```{r}
model.21 <- lm(BP ~ BSA + Weight, data = bloodpress)
summary(model.21)
anova(model.21)
```
```{r}
attach(bloodpress)
scatter3d(BP ~ Weight + BSA)
detach(bloodpress)
```
```{r}
predict(model.1, interval="prediction",
        newdata=data.frame(Weight=92))

predict(model.2, interval="prediction",
        newdata=data.frame(BSA=2))

predict(model.12, interval="prediction",
        newdata=data.frame(Weight=92, BSA=2))
```
# Poverty and teen birth rate (high multicollinearity)

    Load the poverty data and remove the District of Columbia.
    Create a scatterplot matrix of the data.
    Fit a simple linear regression model of PovPct vs Brth15to17.
    Fit a simple linear regression model of PovPct vs Brth18to19.
    Fit a multiple linear regression model of PovPct vs Brth15to17 + Brth18to19.

```{r}
poverty <- read.table("DataL12/poverty.txt", header=T)
poverty <- poverty[poverty$Location!="District_of_Columbia",]
pairs(poverty[,c(2:4)])
```
```{r}
model.1 <- lm(PovPct ~ Brth15to17, data=poverty)
summary(model.1)
```
```{r}
model.2 <- lm(PovPct ~ Brth18to19, data=poverty)
summary(model.2)
```
```{r}
model.12 <- lm(PovPct ~ Brth15to17 + Brth18to19, data=poverty)
summary(model.12)
```
# Blood pressure (high multicollinearity)

    Load the bloodpress data.
    Fit a multiple linear regression model of BP vs Age + Weight + BSA + Dur + Pulse + Stress.
    Use the vif function in the car package to calculate variance inflation factors.
    Fit a multiple linear regression model of Weight vs Age + BSA + Dur + Pulse + Stress and confirm the VIF value for Weight as 1/(1-R2) for this model.
    Fit a multiple linear regression model of BP vs Age + Weight + Dur + Stress.
    Use the vif function in the car package to calculate variance inflation factors.
```{r}
model.1 <- lm(BP ~ Age + Weight + BSA + Dur + Pulse + Stress, data = bloodpress)
summary(model.1)
```
```{r}
# library(car)
vif(model.1)
```
```{r}
model.2 <- lm(Weight ~ Age + BSA + Dur + Pulse + Stress, data=bloodpress)
summary(model.2)
```
```{r}
1/(1-summary(model.2)$r.squared) # 8.417035
```
```{r}
model.3 <- lm(BP ~ Age + Weight + Dur + Stress, data=bloodpress)
summary(model.3)
```
```{r}
vif(model.3)
```
# Allen Cognitive Level study (reducing data-based multicollinearity)

    Load the sampled allentestn23 data.
    Create a scatterplot matrix of the data.
    Calculate the correlation between Vocab and Abstract.
    Fit a multiple linear regression model of ACL vs SDMT + Vocab + Abstract.
    Use the vif function in the car package to calculate variance inflation factors.
    Repeat for the full allentest data.
```{r}
allentestn23 <- read.table("DataL12/allentestn23.csv", sep=",",  header=T)
pairs(allentestn23[,2:5])
```
```{r}
attach(allentestn23)
cor(Vocab, Abstract)
detach(allentestn23)
```
```{r}
model.1 <- lm(ACL ~ SDMT + Vocab + Abstract, data=allentestn23)
summary(model.1)
vif(model.1)
```
```{r}
allentest <- read.table("DataL12/allentest.txt", header=T)

pairs(allentest[,2:5])
```
```{r}
attach(allentest)
cor(Vocab, Abstract)
detach(allentest)
```
```{r}
model.1 <- lm(ACL ~ SDMT + Vocab + Abstract, data=allentest)
summary(model.1)
vif(model.1)
```
# Exercise and immunity (reducing structural multicollinearity)

    Load the exerimmun data.
    Create a scatterplot of igg vs oxygen.
    Calculate an oxygen-squared variable named oxygensq.
    Fit a quadratic regression model of igg vs oxygen + oxygensq.
    Add a quadratic regression line to the scatterplot.
    Use the vif function in the car package to calculate variance inflation factors.
    Create a scatterplot of oxygensq vs oxygen and calculate the correlation.
    Calculate a centered oxygen variable named oxcent and an oxcent-squared variable named oxcentsq.
    Fit a quadratic regression model of igg vs oxcent + oxcentsq.
    Use the vif function in the car package to calculate variance inflation factors.
    Create a scatterplot of igg vs oxcent with the quadratic regression line added.
    Fit a simple linear regression model of igg vs oxcent.
    Confirm the equivalence of the original quadratic and centered quadratic models by transforming the regression parameter estimates.
    Create a residual vs fits plot for the centered quadratic model.
    Create a normal probability plot of the residuals for the centered quadratic model.
    Predict igg for oxygen = 70 using the centered quadratic model.
```{r}
exerimmun <- read.table("DataL12/exerimmun.csv", sep=","  ,header=T)
qplot(data=exerimmun, oxygen, igg)
```
```{r}
exerimmun$oxygensq <- exerimmun$oxygen^2

model.1 <- lm(igg ~ oxygen + oxygensq, data=exerimmun)

qplot(data=exerimmun, x=oxygen, y=igg) +
  geom_line(aes(x=oxygen, y=model.1$fitted.values), col="blue")
```
```{r}
summary(model.1)
vif(model.1)
```
```{r}
qplot(data=exerimmun, oxygen, oxygensq)
```
```{r}
with(data=exerimmun, cor(oxygen, oxygensq))
```
```{r}
exerimmun$oxcent <- exerimmun$oxygen-mean(exerimmun$oxygen)
exerimmun$oxcentsq <- exerimmun$oxcent^2
qplot(data=exerimmun, oxcent, oxcentsq)
```
```{r}
with(data=exerimmun, cor(oxcent, oxcentsq))
```
```{r}
model.2 <- lm(igg ~ oxcent + oxcentsq, data=exerimmun)
summary(model.2)
vif(model.2)
```
```{r}
qplot(data=exerimmun, x=oxygen, y=igg) +
  geom_line(aes(x=oxygen, y=model.2$fitted.values), col="red")
```
```{r}
model.3 <- lm(igg ~ oxcent, data=exerimmun)
summary(model.3)
```
```{r}
with(data=exerimmun, 
coef(model.2)[1]-coef(model.2)[2]*mean(oxygen)+coef(model.2)[3]*mean(oxygen)^2)

with(data=exerimmun,
coef(model.2)[2]-2*coef(model.2)[3]*mean(oxygen))

with(data=exerimmun, coef(model.2)[3] )
```
```{r}
coef(model.1)
```
```{r}
qplot(x=fitted(model.2), y=residuals(model.2)) +
   geom_hline(yintercept = 0)
```
```{r}
qqnorm(residuals(model.2), main="", datax=TRUE)
qqline(residuals(model.2), datax=TRUE)

```
```{r}
with(data=exerimmun, 
predict(model.2, interval="prediction",
        newdata=data.frame(oxcent=70-mean(oxygen), oxcentsq=(70-mean(oxygen))^2))
)
```
df <- data.frame(y = rt(200, df = 5))
p <- ggplot(df, aes(sample = y))
p + stat_qq() + stat_qq_line()

```{r}
df <- data.frame(x = residuals(model.2))
p <- ggplot(df, aes(sample = x))
p + stat_qq() + stat_qq_line()
```

