---
title: "Lesson 8: Categorical Predictors"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
# Birthweight and smoking (2-level categorical predictor, additive model)
https://onlinecourses.science.psu.edu/stat501/node/426/
* Load the birthsmokers data.
* Create a scatterplot matrix of the data
* Fit a multiple linear regression model of Wgt on Gest + Smoke.
* Display scatterplot of Wgt vs Gest with points marked by Smoke and add parallel regression lines representing Smoke=0 and Smoke=1.
* Display regression results and calculate confidence intervals for the regression parameters.
* Display confidence intervals for expected Wgt at Gest=38 (for Smoke=1 and Smoke=0).
* Repeat analysis separately for Smoke=0 and Smoke=1.
* Repeat analysis using (1, -1) coding.

```{r message=FALSE, warning=FALSE}
birthsmokers <- read.csv("DataL8/birthsmokers.csv", header=T)
attach(birthsmokers)
head(birthsmokers)
# str(birthsmokers)
birthsmokers$Smoke <- as.factor(birthsmokers$Smoke)
str(birthsmokers)
pairs(cbind(Wgt, Gest, Smoke), lower.panel = NULL)
# head(birthsmokers)
```
Build linear model as follows :
```{r}
model <- lm(Wgt ~ Gest + Smoke)

plot(x=Gest, y=Wgt, ylim=c(2300, 3700), 
     col=ifelse(Smoke==1, "red", "blue"),
     panel.last = c(lines(sort(Gest[Smoke==0]),
                          fitted(model)[Smoke==0][order(Gest[Smoke==0])],
                          col="blue"),
                    lines(sort(Gest[Smoke==1]),
                          fitted(model)[Smoke==1][order(Gest[Smoke==1])],
                          col="red")))
```
```{r}
summary(model)
```
```{r}
confint(model)
```
```{r}
predict(model, interval="confidence",
        newdata=data.frame(Gest=c(38, 38), Smoke=c(1, 0)))
```
To investigate the 'non-smoker' as subset in the linear model
```{r}
model.0 <- lm(Wgt ~ Gest, subset=Smoke==0) # changed to '0' from "no"
summary(model.0)
```
```{r}
predict(model.0, interval="confidence",
        newdata=data.frame(Gest=38))
```
To investigate the 'smoker' as subset in the linear model
```{r}
model.1 <- lm(Wgt ~ Gest, subset=Smoke==1) # changed to 1 from 'Yes'
summary(model.1)
```
```{r}
predict(model.1, interval="confidence",
        newdata=data.frame(Gest=38))
```
```{r}
Smoke2 <- ifelse(Smoke=="yes", 1, -1)
model.3 <- lm(Wgt ~ Gest + Smoke2)
summary(model.3)
detach(birthsmokers)
```
# Depression treatments (3-level categorical predictor, interaction model)

Load the depression data.
```{r}
depression <- read.table("DataL8/depression.txt", header=T)
attach(depression)
head(depression)
```
Display scatterplot of y (treatment effectiveness) vs age with points marked by treatment.
```{r}
with(data=depression, 
plot(x=age, y=y, col=as.numeric(TRT)),
legend("topleft", col=1:3, pch=1,
       inset=0.02, x.intersp = 1.5, y.intersp = 1.8,
       legend=c("Trt A", "Trt B", "Trt C"))
)
```
Create interaction variables and fit a multiple linear regression model of y on age + x2 + x3 + age.x2 + age.x3.
```{r}
age.x2 <- age*x2
age.x3 <- age*x3

model.1 <- lm(y ~ age + x2 + x3 + age.x2 + age.x3, data = depression) # add data=
summary(model.1)
# using attach () and detach may cause confusions and trouble
```
Add non-parallel regression lines representing each of the three treatments to the scatterplot.
```{r}
with(data=depression, 
plot(x=age, y=y, ylim=c(20, 80), col=as.numeric(TRT),
     panel.last = c(lines(sort(age[TRT=="A"]),
                          fitted(model.1)[TRT=="A"][order(age[TRT=="A"])],
                          col=1),
                    lines(sort(age[TRT=="B"]),
                          fitted(model.1)[TRT=="B"][order(age[TRT=="B"])],
                          col=2),
                    lines(sort(age[TRT=="C"]),
                          fitted(model.1)[TRT=="C"][order(age[TRT=="C"])],
                          col=3))), 
legend("topleft", col=1:3, pch=1, lty=1,
       inset=0.02, x.intersp = 1.5, y.intersp = 1.8,
       legend=c("Trt A", "Trt B", "Trt C"))
)
```
Display a residuals vs fits plot and a normal probability plot of the residuals, and conduct an Anderson-Darling normality test using the nortest package.
```{r}
plot(x=fitted(model.1), y=rstudent(model.1),
     panel.last = abline(h=0, lty=2))

qqnorm(residuals(model.1), main="", datax=TRUE)
qqline(residuals(model.1), datax=TRUE)

library(nortest)
ad.test(residuals(model.1)) # A = 0.40575, p-value = 0.3345
```
Conduct an F-test to see if at least one of x2, x3, age.x2, and age.x3 are useful (i.e., the regression functions differ).
```{r}
anova(model.1)
```
Conduct an F-test to see if at least one of age.x2 and age.x3 are useful (i.e., the regression functions have different slopes).
```{r}
model.2 <- lm(y ~ age, data=depression)
anova(model.2, model.1)
```
```{r}
model.3 <- lm(y ~ age + x2 + x3, data=depression)
anova(model.3, model.1)
detach(depression) # may should not use this anymore
```
https://onlinecourses.science.psu.edu/stat501/node/383/
# Real estate air conditioning (2-level categorical predictor, interaction model, transformations)

* Load the realestate data.
```{r}
realestate <- read.table("DataL8/realestate.txt", header=T)
# attach(realestate)
```
* Create an interaction variable and fit a multiple linear regression model of SalePrice on SqFeet + Air + SqFeet.Air.
```{r}
realestate$SqFeet.Air <- realestate$SqFeet*realestate$Air
model.1 <- lm(SalePrice ~ SqFeet + Air + SqFeet.Air, data=realestate)
summary(model.1)
```
* Display scatterplot of SalePrice vs SqFeet with points marked by Air and add non-parallel regression lines representing Air=0 and Air=1.
```{r}
with(data=realestate, 
plot(x=SqFeet, y=SalePrice, 
     col=ifelse(Air, "red", "blue"),
     panel.last = c(lines(sort(SqFeet[Air==0]),
                          fitted(model.1)[Air==0][order(SqFeet[Air==0])],
                          col="blue"),
                    lines(sort(SqFeet[Air==1]),
                          fitted(model.1)[Air==1][order(SqFeet[Air==1])],
                          col="red")))) 
legend("topleft", col=c("blue", "red"), pch=1, lty=1, inset=0.02,
       legend=c("No air conditioning", "Air conditioning"))
# this is for 'with'
```
* Display residual plot with fitted (predicted) values on the horizontal axis.
```{r}
plot(x=fitted(model.1), y=residuals(model.1),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))

```

* Create log(SalePrice), log(SqFeet), and log(SqFeet).Air variables and fit a multiple linear regression model of log(SalePrice) on log(SqFeet) + Air + log(SqFeet).Air.
```{r}
realestate$lnSalePrice <- log(realestate$SalePrice)
realestate$lnSqFeet <- log(realestate$SqFeet)
realestate$lnSqFeet.Air <- realestate$lnSqFeet*realestate$Air

model.2 <- lm(lnSalePrice ~ lnSqFeet + Air + lnSqFeet.Air, data=realestate)

```
* Display scatterplot of log(SalePrice) vs log(SqFeet) with points marked by Air and add non-parallel regression lines representing Air=0 and Air=1.
```{r}
with(data=realestate, 
plot(x=lnSqFeet, y=lnSalePrice, 
     col=ifelse(Air, "red", "blue"),
     panel.last = c(lines(sort(lnSqFeet[Air==0]),
                          fitted(model.2)[Air==0][order(lnSqFeet[Air==0])],
                          col="blue"),
                    lines(sort(lnSqFeet[Air==1]),
                          fitted(model.2)[Air==1][order(lnSqFeet[Air==1])],
                          col="red"))))
legend("topleft", col=c("blue", "red"), pch=1, lty=1, inset=0.02,
       legend=c("No air conditioning", "Air conditioning"))

```
* Display residual plot with fitted (predicted) values on the horizontal axis.
```{r}
plot(x=fitted(model.2), y=residuals(model.2),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))

```
# Hospital infection risk (4-level categorical predictor, additive model)

Load the infectionrisk data and select observations with Stay <= 14.
Create indicator variables for regions.
```{r}
infectionrisk <- read.table("DataL8/infectionrisk.txt", header=T)
infectionrisk <- infectionrisk[infectionrisk$Stay<=14,]
# attach(infectionrisk)

infectionrisk$i1 <- ifelse(infectionrisk$Region==1,1,0) # NE
infectionrisk$i2 <- ifelse(infectionrisk$Region==2,1,0) # NC
infectionrisk$i3 <- ifelse(infectionrisk$Region==3,1,0) # S
infectionrisk$i4 <- ifelse(infectionrisk$Region==4,1,0) # W

```

Fit a multiple linear regression model of InfctRsk on Stay + Xray + i2 + i3 + i4.
```{r}
head(infectionrisk)
model.1 <- lm(InfctRsk ~ Stay + Xray + i2 + i3 + i4, data = infectionrisk)
summary(model.1)
```

Conduct an F-test to see if at least one of i2, i3, and i4 are useful (conclusion: the regression functions differ by region).
```{r}
model.2 <- lm(InfctRsk ~ Stay + Xray, data = infectionrisk)
anova(model.2, model.1)
```
Conduct an F-test to see if at least one of i2 and i3 are useful (conclusion: only the west region differs).
```{r}
model.3 <- lm(InfctRsk ~ Stay + Xray + i4, data=infectionrisk)
anova(model.3, model.1)
```


