---
title: "Lesson 13: Weighted Least Squares & Robust Regression"
output: html_notebook
---
https://onlinecourses.science.psu.edu/stat501/node/431/
# Galton peas (nonconstant variance and weighted least squares)
Load the galton data.
Fit an ordinary least squares (OLS) simple linear regression model of Progeny vs Parent.
```{r}
galton <- read.table("DataL13/galton.txt", header=T)
head(galton)
model.1 <- lm(Progeny ~ Parent, data=galton)
summary(model.1)
```
Fit a weighted least squares (WLS) model using weights = 1/SD2.
```{r}
model.2 <- lm(Progeny ~ Parent, weights=1/SD^2, data=galton)
summary(model.2)
```
Create a scatterplot of the data with a regression line for each model.
```{r}
with(data=galton, 
plot(x=Parent, y=Progeny, ylim=c(0.158,0.174),
     panel.last = c(lines(sort(Parent), fitted(model.1)[order(Parent)], col="blue"),
                    lines(sort(Parent), fitted(model.2)[order(Parent)], col="red")))
)
legend("topleft", col=c("blue","red"), lty=1,
       inset=0.02, legend=c("OLS", "WLS"))

```
# Computer-assisted learning (nonconstant variance and weighted least squares)
Load the ca_learning data.
Create a scatterplot of the data.
```{r}
ca_learning <- read.table("DataL13/ca_learning_new.csv", sep=",", header=T)

with(data=ca_learning, plot(x=num.responses, y=cost))
```
Fit an OLS model.
```{r}
model.1 <- lm(cost ~ num.responses, data=ca_learning)
summary(model.1)
```
Plot the OLS residuals vs num.responses.
```{r}
with(data=ca_learning,
plot(num.responses, residuals(model.1)))
```
Plot the absolute OLS residuals vs num.responses.
```{r}
with(data=ca_learning,
plot(num.responses, abs(residuals(model.1))))
```
Calculate fitted values from a regression of absolute residuals vs num.responses.
Fit a WLS model using weights = 1/(fitted values)2.
```{r}
wts <- 1/fitted(lm(abs(residuals(model.1)) ~ num.responses, data=ca_learning))^2

model.2 <- lm(cost ~ num.responses, weights=wts, data=ca_learning)
summary(model.2)
```
Create a scatterplot of the data with a regression line for each model.
```{r}
with(data=ca_learning,
plot(x=num.responses, y=cost, ylim=c(50,95),
     panel.last = c(lines(sort(num.responses), fitted(model.1)[order(num.responses)], col="blue"),
                    lines(sort(num.responses), fitted(model.2)[order(num.responses)], col="red"))))

legend("topleft", col=c("blue","red"), lty=1,
       inset=0.02, legend=c("OLS", "WLS"))
```
Plot the WLS standardized residuals vs num.responses.
```{r}
with(data=ca_learning,
plot(num.responses, rstandard(model.2)))
```
# Market share (nonconstant variance and weighted least squares)
Load the marketshare data.
Fit an OLS model.
```{r}
marketshare <- read.table("DataL13/market_share.csv", sep=",",  header=T)
head(marketshare)
model.1 <- lm(Market.share ~ Price + Discount + Promotion, data=marketshare)
summary(model.1)
```
Plot the OLS residuals vs fitted values with points marked by Discount.
```{r}
plot(fitted(model.1), residuals(model.1), col=marketshare$Discount+1)
```
Use the tapply function to calculate the residual variance for Discount=0 and Discount=1.
```{r}
vars <- tapply(residuals(model.1), marketshare$Discount, var)
vars
```
Fit a WLS model using weights = 1/variance for Discount=0 and Discount=1.
```{r}
attach(marketshare)
wts <- Discount/vars[2] + (1-Discount)/vars[1]
model.2 <- lm(Market.share ~ Price + Discount + Promotion, weights=wts)
summary(model.2)
```
Plot the WLS standardized residuals vs fitted values.
```{r}
plot(fitted(model.2), rstandard(model.2), col=marketshare$Discount+1)
detach(marketshare)
```
# Home price (nonconstant variance and weighted least squares)

Load the realestate data.
Calculate log transformations of the variables.
Fit an OLS model.
```{r}
realestate <- read.table("DataL13/home_price.csv", sep=",",  header=T)

head(realestate)
realestate <- realestate[,-1] # remove 'ID'

colnames(realestate) <- c("SalePrice", "SqFeet", "Lot")
attach(realestate)
logY <- log(SalePrice)
logX1 <- log(SqFeet)
logX2 <- log(Lot)

model.1 <- lm(logY ~ logX1 + logX2)
summary(model.1)
```
Plot the OLS residuals vs fitted values.
```{r}
plot(fitted(model.1), residuals(model.1))
```
Calculate fitted values from a regression of absolute residuals vs fitted values.
Fit a WLS model using weights = 1/(fitted values)2.
```{r}
wts <- 1/fitted(lm(abs(residuals(model.1)) ~ fitted(model.1)))^2
model.2 <- lm(logY ~ logX1 + logX2, weights=wts)
summary(model.2)
```
Plot the WLS standardized residuals vs fitted values.
```{r}
plot(fitted(model.2), rstandard(model.2))

detach(realestate)
```
