---
title: "Lesson 6: MLR Model Evaluation"
output: html_notebook
---
Heart attacks in rabbits
https://newonlinecourses.science.psu.edu/stat501/node/424/
    Load the coolhearts data.
    Fit a multiple linear regression model of Infarc on Area, X2 (early cooling), and X3 (late cooling).
    Display model results.
    Create a scatterplot of the data with points marked by group and three lines representing the fitted regression equation for each group.
```{r}
coolhearts <- read.csv("DataL6/coolhearts.csv", header=T)
colnames(coolhearts) <- c("Inf", "Area", "Group", "X2", "X3")
# str(coolhearts)
coolhearts$X2 <- as.numeric(coolhearts$X2)
coolhearts$X3 <- as.numeric(coolhearts$X3)
# attach(coolhearts)
# head(coolhearts)
# tail(coolhearts)
model.1 <- lm( Inf.~ Area + X2 + X3, data=coolhearts) # canno have 'inf' ?
summary(model.1)
```
```{r message=FALSE}
attach(coolhearts)
plot(Area, Inf., type="n", ylim=c(-0.2, 1),
     xlab="Size of area at risk (grams)",
     ylab="Size of infarcted area (grams)")
for (i in 1:32) points(Area[i], Inf.[i], pch=Group[i], col=Group[i])
for (i in 1:3) lines(Area[Group==i], fitted(model.1)[Group==i], lty=i, col=i)
legend("topleft", legend=c("no cooling", 
                           "late cooling",
                           "early cooling"),
       col=3:1, pch=3:1, inset=0.02)
detach(coolhearts)
```
Student heights and GPAs

    Load the heightgpa data.
    Fit a simple linear regression model of gpa on height.
    Create a scatterplot of the data with a fitted simple linear regression line and a horizontal line at the mean of gpa.
    Calculate SSE for the full and reduced models.
    Calculate the overall F statistic by hand.
    Find the p-value for the overall F statistic.
    Display the overall F statistic and p-value in the model results.
```{r}
heightgpa <- read.table("DataL6/heightgpa.txt", header=T)
attach(heightgpa)

model <- lm(gpa ~ height)

plot(height,gpa,xlab="Height (inches)",ylab="Grade Point Average",
     panel.last = c(lines(sort(height), fitted(model)[order(height)]),
                    abline(h=mean(gpa), lty=2)))
```
```{r}
sum(residuals(model)^2) # SSE_F = 9.705507
mean(gpa) # 2.971714
sum((gpa-mean(gpa))^2) # SSE_R = 9.733097
((9.733097-9.705507)/1) / (9.705507/33) # overall F-statistic = 0.09380963
pf(0.09380963, 1, 33, lower.tail=F) # p-value = 0.7613129
```
```{r}
summary(model) # F-statistic: 0.09381 on 1 and 33 DF,  p-value: 0.7613

detach(heightgpa)
```
Skin cancer mortality

    Load the skincancer data.
    Fit a simple linear regression model of Mort on Lat.
    Create a scatterplot of the data with a fitted simple linear regression line and a horizontal line at the mean of Mort.
    Calculate SSE for the full and reduced models.
    Calculate the overall F statistic by hand.
    Find the p-value for the overall F statistic.
    Display the overall F statistic and p-value in the model results.
```{r}
skincancer <- read.table("DataL6/skincancer.txt", header=T)
attach(skincancer)

model <- lm(Mort ~ Lat)

plot(Lat,Mort,xlab="Latitude (at center of state)",
     ylab="Mortality (deaths per 10 million)",
     panel.last = c(lines(sort(Lat), fitted(model)[order(Lat)]),
                    abline(h=mean(Mort), lty=2)))

```
```{r}
sum(residuals(model)^2) # SSE_F = 17173.07
mean(Mort) # 152.8776
sum((Mort-mean(Mort))^2) # SSE_R = 53637.27
((53637.27-17173.07)/1) / (17173.07/47) # overall F-statistic = 99.7968
pf(99.7968, 1, 47, lower.tail=F) # p-value = 3.309471e-13
```
```{r}
summary(model) # F-statistic:  99.8 on 1 and 47 DF,  p-value: 3.309e-13

detach(skincancer)
```
Alcoholism and muscle strength

    Load the alcoholarm data.
    Fit a simple linear regression model of strength on alcohol.
    Calculate SSE for the full and reduced models.
    Calculate the overall F statistic by hand.
    Display the overall F statistic and p-value in the model results and in the anova table.
```{r}
alcoholarm <- read.table("DataL6/alcoholarm.txt", header=T)
attach(alcoholarm)

model <- lm(strength ~ alcohol)
sum((strength-mean(strength))^2) # SSE_R = 1224.315
sum(residuals(model)^2) # SSE_F = 720.2749
((1224.315-720.2749)/1) / (720.2749/48) # F = 33.58985
summary(model) # F-statistic: 33.59 on 1 and 48 DF,  p-value: 5.136e-07
```
```{r}
anova(model)

detach(alcoholarm)
```
Allen cognitive level study

    Load the allentest data.
    Calculate SSTO.
    Fit SLR model of ACL on Vocab and display anova table (with sequential sums of squares).
    Fit MLR model of ACL on Vocab and SDMT and display anova table (with sequential sums of squares).
    Calculate SSR(Vocab, SDMT) by hand using sequential sums of squares.
    Fit SLR model of ACL on SDMT and display anova table (with sequential sums of squares).
    Fit MLR model of ACL on SDMT and Vocab and display anova table (with sequential sums of squares).
    Calculate SSR(Vocab, SDMT) by hand using sequential sums of squares.
    Fit MLR model of ACL on SDMT, Vocab, and Abstract and display anova table (with sequential sums of squares).
    Calculate SSR(Vocab, Abstract | SDMT) by hand using sequential sums of squares.
```{r}
allentest <- read.table("DataL6/allentest.txt", header=T)
attach(allentest)
head(allentest)

sum((ACL-mean(ACL))^2) # SSTO = 43.04957

model.1 <- lm(ACL ~ Vocab)
anova(model.1)
```
```{r}
model.13 <- lm(ACL ~ Vocab + SDMT)
anova(model.13) # Sequential (type I) SS
```
Calculate by hand: SSR(Vocab, SDMT) = 2.6906 + 9.0872 = 11.7778

```{r}
model.3 <- lm(ACL ~ SDMT)
anova(model.3)
```
```{r}
model.31 <- lm(ACL ~ SDMT + Vocab)
anova(model.31) # Sequential (type I) SS
```
Calculate by hand: SSR(Vocab, SDMT) = 11.6799 + 0.0979 = 11.7778
```{r}
model.312 <- lm(ACL ~ SDMT + Vocab + Abstract)
anova(model.312) # Sequential (type I) SS
detach(allentest)
```
Calculate by hand: SSR(Vocab, Abstract | SDMT) = 0.0979 + 0.5230 = 0.6209

Heart attacks in rabbits (revisited)

    Load the coolhearts data.
    Fit a multiple linear regression model of Infarc on Area, X2 (early cooling), and X3 (late cooling).
    Test all slope parameters equal 0.
        Display the overall F statistic and p-value in the model results.
    Test one slope parameter is 0.
        Use the Anova function from the car package to display F-statistic in anova table using adjusted (type III) sums of squares.
        Or (easier), use t-test from model results.
    Test a subset of slope parameters is 0.
        Fit full model (with Area, X2 , and X3) and reduced model (Area only), and calculate general linear F-statistic.
        Or, use the anova function with full model to display anova table with sequential (type I) sums of squares, and calculate partial F-statistic.
        Or (easier), use the anova function with full and reduced models to display F-statistic and p-value directly.
    Calculate partial R-squared for (X2, X3 | Area).
```{r message=FALSE}
coolhearts <- read.csv("DataL6/coolhearts.csv", header=T)
colnames(coolhearts) <- c("Infarc", "Area", "Group", "X2", "X3")
attach(coolhearts)
coolhearts$X2 <- as.numeric(coolhearts$X2)
coolhearts$X3 <- as.numeric(coolhearts$X3)

model.1 <- lm(Infarc ~ Area + X2 + X3)
summary(model.1)
```
```{r}
anova(model.1) # Sequential (type I) SS
```
```{r}
library(alr3)
Anova(model.1, type="III") # Adjusted (type III) SS
```
 Hand calculation: F for Area = (0.63742/1) / (0.54491/28) = 32.75359
 
```{r}
pf(32.75359, 1, 28, lower.tail=F) # 3.865451e-06
5.723^2 # 32.75273 (t-statistic squared = F-statistic) 
```
```{r}
model.2 <- lm(Infarc ~ Area)
anova(model.2)
((0.87926-0.54491)/(30-28)) / (0.54491/28) # General linear F-stat = 8.590226
((0.31453+0.01981)/2) / (0.54491/28) # Partial F-stat = 8.589969
pf(8.59, 2, 28, lower.tail=F) # 0.001233006
```
```{r}
anova(model.2, model.1)
```
```{r}
0.31453+0.01981 # SSR(X2, X3 | Area) = 0.33434
# SSE(Area) = 0.87926
0.33434 / 0.87926 # Partial R-squared (X2, X3 | Area) = 0.3802516

detach(coolhearts)
```
 


