---
title: "Lesson 8: Categorical Predictors"
output: html_notebook
---
Birthweight and smoking (2-level categorical predictor, additive model)
https://onlinecourses.science.psu.edu/stat501/node/426/

    Load the birthsmokers data.
    Create a scatterplot matrix of the data
    Fit a multiple linear regression model of Wgt on Gest + Smoke.
    Display scatterplot of Wgt vs Gest with points marked by Smoke and add parallel regression lines representing Smoke=0 and Smoke=1.
    Display regression results and calculate confidence intervals for the regression parameters.
    Display confidence intervals for expected Wgt at Gest=38 (for Smoke=1 and Smoke=0).
    Repeat analysis separately for Smoke=0 and Smoke=1.
    Repeat analysis using (1, -1) coding.

```{r message=FALSE, warning=FALSE}
birthsmokers <- read.csv("DataL8/birthsmokers.csv", header=T)
attach(birthsmokers)
# head(birthsmokers)
# str(birthsmokers)
birthsmokers$Smoke <- as.factor(birthsmokers$Smoke)
str(birthsmokers)
pairs(cbind(Wgt, Gest, Smoke))
head(birthsmokers)
```
```{r}
model <- lm(Wgt ~ Gest + Smoke)

plot(x=Gest, y=Wgt, ylim=c(2300, 3700), 
     col=ifelse(Smoke==1, "red", "blue"),
     panel.last = c(lines(sort(Gest[Smoke==0]),
                          fitted(model)[Smoke==0][order(Gest[Smoke==0])],
                          col="blue"),
                    lines(sort(Gest[Smoke==1]),
                          fitted(model)[Smoke==1][order(Gest[Smoke==1])],
                          col="red")))
```
```{r}
summary(model)
```
```{r}
confint(model)
```
```{r}
predict(model, interval="confidence",
        newdata=data.frame(Gest=c(38, 38), Smoke=c(1, 0)))
```
```{r}
model.0 <- lm(Wgt ~ Gest, subset=Smoke==0) # changed to '0' from "no"
summary(model.0)
```
```{r}
predict(model.0, interval="confidence",
        newdata=data.frame(Gest=38))
```
```{r}
model.1 <- lm(Wgt ~ Gest, subset=Smoke==1) # changed to 1 from 'Yes'
summary(model.1)
```
```{r}
predict(model.1, interval="confidence",
        newdata=data.frame(Gest=38))
```
```{r}
Smoke2 <- ifelse(Smoke=="yes", 1, -1)
model.3 <- lm(Wgt ~ Gest + Smoke2)
summary(model.3)
detach(birthsmokers)
```


