---
title: "Lesson 11: Influential Points"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
https://onlinecourses.science.psu.edu/stat501/node/429/
```{r}
library(knitr)
```
# Influence 1 (no influential points)
Load the influence1 data.
Create a scatterplot of the data.
```{r message=FALSE}
# getwd()
influence1 <- read.csv("DataL11/influence1.csv", header=T)
library(ggplot2)
attach(influence1)
# head(influence1)
qplot(data=influence1, x=x, y=y)

detach(influence1)
```
# Influence 2 (outlier, low leverage, not influential)
Load the influence2 data.
Create a scatterplot of the data.
```{r message=FALSE}
influence2 <- read.csv("DataL11/influence2.csv", header=T)
attach(influence2)
# head(influence2)
qplot(data= influence2, x=x, y=y)
```
Fit a simple linear regression model to all the data.
```{r}
model.1 <- lm(y ~ x, data = influence2)
summary(model.1)
```
Fit a simple linear regression model to the data excluding observation #21.
```{r}
model.2 <- lm(y ~ x, subset = 1:20) # exclude obs #21
summary(model.2)
```
Add regression lines to the scatterplot, one for each model.
## clean the data if not working...!!!
```{r}
plot(x=x, y=y, col=ifelse(Row<=20, "blue", "red"),
     panel.last = c(lines(sort(x), fitted(model.1)[order(x)], col="red"),
                    lines(sort(x), fitted(model.2)[order(x)],
                          col="red", lty=2)))
legend("topleft", col="red", lty=c(1,2),
       inset=0.02, legend=c("Red point included", "Red point excluded"))
```
Calculate leverages, standardized residuals, studentized residuals, DFFITS, Cook's distances. 
```{r}
lev <- hatvalues(model.1)
round(lev, 6)
plot(lev)
```
```{r}
sum(lev) # 2
```
```{r}
sta <- rstandard(model.1)
round(sta, 6)
plot(sta)
```
```{r}
stu <- rstudent(model.1)
round(stu, 6)
plot(stu)
```
```{r}
dffit <- dffits(model.1)
round(dffit, 6)
plot(dffit)
```
```{r}
cook <- cooks.distance(model.1)
round(cook, 6)
plot(cook)
detach(influence2)
```
# Influence 3 (high leverage, not an outlier, not influential)
Load the influence3 data.
Create a scatterplot of the data.
```{r}
influence3 <- read.csv("DataL11/influence3.csv", header=T)
attach(influence3)
plot(x, y)
```
Fit a simple linear regression model to all the data.
Fit a simple linear regression model to the data excluding observation #21.
```{r}
model.1 <- lm(y ~ x)
summary(model.1)
```
```{r}
model.2 <- lm(y ~ x, subset=1:20) # exclude obs #21
summary(model.2)
```
Add regression lines to the scatterplot, one for each model.
```{r}
plot(x=x, y=y, col=ifelse(Row<=20, "blue", "red"),
     panel.last = c(lines(sort(x), fitted(model.1)[order(x)], col="red"),
                    lines(sort(x[-21]), fitted(model.2)[order(x[-21])],
                          col="red", lty=2)))
legend("topleft", col="red", lty=c(1,2),
       inset=0.02, legend=c("Red point included", "Red point excluded"))

```
Calculate leverages, DFFITS, Cook's distances.
```{r}
lev <- hatvalues(model.1)
round(lev, 6)
```
```{r}
dffit <- dffits(model.1)
round(dffit, 6)
```
```{r}
cook <- cooks.distance(model.1)
round(cook, 6)
plot(cook)
detach(influence3)
```
# Influence 4 (outlier, high leverage, influential)

Load the influence4 data.
Create a scatterplot of the data.
```{r message=FALSE}
influence4 <- read.csv("DataL11/influence4.csv", header=T)
attach(influence4)

qplot(x, y)
```
Fit a simple linear regression model to all the data.
Fit a simple linear regression model to the data excluding observation #21.
```{r}
model.1 <- lm(y ~ x)
summary(model.1)
```
```{r}
model.2 <- lm(y ~ x, subset=1:20) # exclude obs #21
summary(model.2)
```
Add regression lines to the scatterplot, one for each model.
```{r}
qplot(data=influence4, x=x, y=y) +
  geom_line(aes(x=x, y=model.1$fitted.values, colour="whole data")) +
  geom_line(data=influence4[-21, ], aes(x=x, y=model.2$fitted.values, colour="no outlier")) +
  xlim(c(0, 16))

# influence4[-21, ]            
```
```{r}
plot(x=x, y=y, col=ifelse(Row<=20, "blue", "red"),
     panel.last = c(lines(sort(x), fitted(model.1)[order(x)], col="red"),
                    lines(sort(x[-21]), fitted(model.2)[order(x[-21])],
                          col="red", lty=2)))
legend("topleft", col="red", lty=c(1,2),
       inset=0.02, legend=c("Red point included", "Red point excluded"))

```
Calculate leverages, DFFITS, Cook's distances.
You can clearly see the slope for model.1 is `r round(model.1$coefficients["x"], 4)` while the slope for model.2 is
`r round(model.2$coefficients["x"], 4)`

## Potentially influential observations 
```{r}
summary(influence.measures(model.1))
```

```{r}
lev <- hatvalues(model.1)
round(lev, 6)
```
```{r}
sum(lev) # 2
```
```{r}
dffit <- dffits(model.1)
round(dffit, 6)
```
```{r}
cook <- cooks.distance(model.1)
round(cook, 6)
plot(cook)
detach(influence4)
```
# Foot length and height (outlier, high leverage, influential)

Load the height_foot data.
Create a scatterplot of the data.
```{r}
heightfoot <- read.table("DataL11/height_foot.txt", header=T)
attach(heightfoot)

qplot(height, foot)
```
Fit a simple linear regression model to all the data.
Fit a simple linear regression model to the data excluding observation #28.
```{r}
model.1 <- lm(foot ~ height)
summary(model.1)
```
```{r}
which(height>80) # 28
model.2 <- lm(foot ~ height, subset=(1:33)[-28]) # exclude obs #28
summary(model.2)
```
Calculate DFFITS and Cook's distance for obs #28.
```{r}
dffit <- dffits(model.1)
dffit[28] # -3.200223
```
```{r}
cook <- cooks.distance(model.1)
cook[28] # 3.274466

detach(heightfoot)
```
# Hospital infection risk (two outliers, high leverages)

Load the infection risk data.
Fit a simple linear regression model to all the data.
```{r}
infectionrisk <- read.csv("DataL11/hospital_infct_03.csv", header=T)
attach(infectionrisk)
head(infectionrisk)
model <- lm(InfctRsk ~ Stay)
summary(model)
```
Create a scatterplot of the data and add the regression line.
```{r}
qplot(data=infectionrisk, x=Stay, y=InfctRsk) +
  geom_line(aes(x=Stay, y= model$fitted.values), col='blue')

# fitted(model)
# x
# names(model)
# model$fitted.values
```
Display influence measures for influential points, including DFFITS, Cook's distances, and leverages (hat).
```{r}
summary(influence.measures(model)) # summarize dffit, cook's dist and hat value
detach(infectionrisk)
```
