---
title: "Lesson 3: SLR Estimation & Prediction"
output: html_notebook
---
# Try to use tidyverse in future examples
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

# Student heights and weights
Load the heightweight data.
Fit a simple linear regression model with y = wt and x = ht.
```{r}
heightweight <- read.table("DataL3/student_height_weight.txt", header=T)
attach(heightweight)
head(heightweight)
model <- lm(wt ~ ht)
```
Display a scatterplot of the data with the simple linear regression line and a horizontal line at the mean weight.
```{r}
# plot(x=ht, y=wt,
#      panel.last = c(lines(sort(ht), fitted(model)[order(ht)]),
#                     abline(h=mean(wt))))
ggplot(data=heightweight, aes(x=ht, y=wt)) +
  geom_point()+
  geom_line(aes(x=ht, y=fitted(model)), col='blue')+
  geom_hline(yintercept = mean(wt))
```
```{r}
mean(wt) # 158.8
```

Use the model to predict weight for height = 64.
```{r}
predict(model, newdata=data.frame(ht=64))

detach(heightweight)
```
# Skin cancer mortality
Load the skin cancer data.
Fit a simple linear regression model with y = Mort and x = Lat.
```{r}
skincancer <- read.table("DataL3/skincancer.txt", header=T)
attach(skincancer)
model <- lm(Mort ~ Lat)
head(skincancer)
```
Display a scatterplot of the data with the simple linear regression line.
```{r}
# plot(x=Lat, y=Mort,
#      xlab="Latitude (at center of state)", ylab="Mortality (deaths per 10 million)",
#      main="Skin Cancer Mortality versus State Latitude",
#      panel.last = lines(sort(Lat), fitted(model)[order(Lat)]))

ggplot(data=skincancer, aes(x=Lat, y=Mort)) +
  geom_point()+
  geom_line(aes(x=Lat, y=fitted(model)), col='blue')+
  xlab("Latitude (at center of state)") +
  ylab("Mortality (deaths per 10 million)")+
  ggtitle("Skin Cancer Mortality versus State Latitude")
```
Use the model to calculate 95% confidence intervals for E(Mort) at Lat = 40 and 28.
```{r}
predict(model, interval="confidence", se.fit=T,
        newdata=data.frame(Lat=c(40, 28)))
```
Calculate mean(Lat).
```{r}
mean(Lat) # 39.53265
```
Use the model to calculate 95% prediction intervals for Mort at Lat = 40.
```{r}
predict(model, interval="prediction",
        newdata=data.frame(Lat=40))
```
Display a scatterplot of the data with the simple linear regression line, confidence interval bounds, and prediction interval bounds.
```{r}
# plot(x=Lat, y=Mort,
#      xlab="Latitude (at center of state)", ylab="Mortality (deaths per 10 million)",
#      ylim=c(60, 260),
#      panel.last = c(lines(sort(Lat), fitted(model)[order(Lat)]),
#                     lines(sort(Lat), 
#                           predict(model, 
#                                   interval="confidence")[order(Lat), 2], col="green"),
#                     lines(sort(Lat), 
#                           predict(model, 
#                                   interval="confidence")[order(Lat), 3], col="green"),
#                     lines(sort(Lat), 
#                           predict(model, 
#                                   interval="prediction")[order(Lat), 2], col="purple"),
#                     lines(sort(Lat), 
#                           predict(model, 
#                                   interval="prediction")[order(Lat), 3], col="purple")))

ggplot(data=skincancer, aes(x=Lat, y=Mort)) +
  geom_point()+
  geom_line(aes(x=Lat, y=fitted(model)), col='blue')+
  xlab("Latitude (at center of state)") +
  ylab("Mortality (deaths per 10 million)")+
  ylim(c(60, 260))+
  ggtitle("Skin Cancer Mortality versus State Latitude")+
  geom_line(aes(x=Lat, y=predict(model, interval = "confidence")[, 2]), col='green')+
  geom_line(aes(x=Lat, y=predict(model, interval = "confidence")[, 3]), col='green')+
  geom_line(aes(x=Lat, y=predict(model, interval = "prediction")[, 2]), col='purple')+
  geom_line(aes(x=Lat, y=predict(model, interval = "prediction")[, 3]), col='purple')

detach(skincancer)
```
# Hospital infection risk

Load the infectionrisk data.
Select only hospitals in regions 1 or 2.
Display a scatterplot of Stay versus InfctRsk.
```{r}
infectionrisk <- read.table("DataL3/infectionrisk.txt", header=T)
infectionrisk <- infectionrisk[infectionrisk$Region==1 | infectionrisk$Region==2, ]
attach(infectionrisk)
qplot(x=Stay, y=InfctRsk)+
  geom_smooth()
detach(infectionrisk)
```
Select only hospitals with Stay < 16 (i.e., remove the two hospitals with extreme values of Stay).
```{r}
infectionrisk <- infectionrisk[infectionrisk$Stay<16, ]
attach(infectionrisk)
qplot(x=Stay, y=InfctRsk)
```
Fit a simple linear regression model with y = InfctRsk and x = Stay.
Use the model to calculate 95% confidence intervals for E(InfctRsk) at Stay = 10.
```{r}
model <- lm(InfctRsk ~ Stay)

predict(model, interval="confidence",
        newdata=data.frame(Stay=10))
```
Use the model to calculate 95% prediction intervals for InfctRsk at Stay = 10.
```{r}
predict(model, interval="prediction",
        newdata=data.frame(Stay=10))
```
Display a scatterplot of the data with the simple linear regression line, confidence interval bounds, and prediction interval bounds.
```{r}
plot(x=Stay, y=InfctRsk,
     ylim=c(0, 9),
     panel.last = c(lines(sort(Stay), fitted(model)[order(Stay)]),
                    lines(sort(Stay), 
                          predict(model, 
                                  interval="confidence")[order(Stay), 2], col="green"),
                    lines(sort(Stay), 
                          predict(model, 
                                  interval="confidence")[order(Stay), 3], col="green"),
                    lines(sort(Stay), 
                          predict(model, 
                                  interval="prediction")[order(Stay), 2], col="purple"),
                    lines(sort(Stay), 
                          predict(model, 
                                  interval="prediction")[order(Stay), 3], col="purple")))

detach(infectionrisk)
```
* We can make the following observations:

For the interval given under 95% CI, we say with 95% confidence we can estimate that in hospitals in which the average length of stay is 10 days, the mean infection risk is between 4.25921 and 4.79849.
For the interval given under 95% PI, we say with 95% confidence that for any future hospital where the average length of stay is 10 days, the infection risk is between 2.45891 and 6.59878.
The value under Fit is calculated as ^y=???1.160+0.5689(10)=4.529

The value under SE Fit is the standard error of ^y and it measures the accuracy of ^y as an estimate of E(Y ).
Since df = n - 2 = 58 - 2 = 56, the multiplier for 95% confidence is 2.00324. The 95% CI for E(Y) is calculated as 4.52885±(2.00324×0.134602)=4.52885±0.26964=(4.259,4.798)

Since S = sqrt(MSE) = 1.02449, the 95% PI is calculated as 4.52885±(2.00324×sqrt(1.024492+0.1346022))=4.52885±0.20699=(2.459,6.599). 
