---
title: "Lesson 2: SLR Model Evaluation"
output: html_notebook
---
Update : 8/26/2018

# E2.1 Skin cancer

* Load the skin cancer data.
* Fit a simple linear regression model with y = Mort and x = Lat.
```{r message=FALSE, warning=FALSE}
skincancer <- read.table("Data2/skincancer.txt", header=T)
attach(skincancer)

model <- lm(Mort ~ Lat)
```
* Display a scatterplot of the data with the simple linear regression line.
```{r}
plot(x=Lat, y=Mort,
     xlab="Latitude (at center of state)", ylab="Mortality (deaths per 10 million)",
     panel.last = lines(sort(Lat), fitted(model)[order(Lat)]))
```
* Display model results.
```{r}
summary(model)
```
* Calculate confidence intervals for the model parameters (regression coefficients).
```{r}
confint(model, level=0.95)
```
```{r}
detach(skincancer)
```
# E2.2 Cord blood lead concentration [data missing]

Load the cord blood lead concentration data.
Fit a simple linear regression model with y = Cord and x = Sold.
```{r}
cordblood <- read.table("Data2/leadcord.txt", header=T)
attach(cordblood)
head(cordblood)
```
Display a scatterplot of the data with the simple linear regression line.
Display model results.
```{r}
model <- lm(Cord ~ Sold)

plot(x=Sold, y=Cord,
     xlab="Monthly gasoline lead sales (metric tons)",
     ylab="Mean cord blood lead concentration",
     panel.last = lines(sort(Sold), fitted(model)[order(Sold)]))
```
```{r}
summary(model)
```
Calculate confidence intervals for the model parameters (regression coefficients).
```{r}
confint(model, level=0.95)
detach(cordblood)
```
# E2.3 Skin cancer
Load the skin cancer data.
Fit a simple linear regression model with y = Mort and x = Lat.
Display analysis of variance table.
```{r}
skincancer <- read.table("Data2/skincancer.txt", header=T)
attach(skincancer)
head(skincancer)
```
```{r}
model <- lm(Mort ~ Lat)
anova(model)
detach(skincancer)
```
* The degrees of freedom associated with SSR will always be 1 for the simple linear regression model. The degrees of freedom associated with SSTO is n-1 = 49-1 = 48. The degrees of freedom associated with SSE is n-2 = 49-2 = 47. And the degrees of freedom add up: 1 + 47 = 48.
The sums of squares add up: SSTO = SSR + SSE. That is, here: 53637 = 36464 + 17173.

# E2.4 Height and grade point average
Load the height and grade point average data.
```{r}
heightgpa <- read.table("Data2/heightgpa.txt", header=T)
head(heightgpa)
attach(heightgpa)
```
Fit a simple linear regression model with y = gpa and x = height.
Display a scatterplot of the data with the simple linear regression line.
```{r}
model <- lm(gpa ~ height)

plot(x=height, y=gpa,
     xlab="Height (inches)", ylab="Grade Point Average",
     panel.last = lines(sort(height), fitted(model)[order(height)]))
```
```{r}
summary(model)
```
Display model results.
Display analysis of variance table.
```{r}
anova(model)
detach(heightgpa)
```
# Sprinters

The F-test is only appropriate for testing that the slope differs from 0 (??1 ??? 0).
Use the t-test to test that the slope is positive (??1 > 0) or negative (??1 < 0). Remember, though, that you will have to divide the P-value that Minitab reports by 2 to get the appropriate P-value.

Load the sprinters data.
Fit a simple linear regression model with y = Men200m and x = Year.
Display a scatterplot of the data with the simple linear regression line.
```{r}
sprinters <- read.table("Data2/mens200m.txt", header=T)
attach(sprinters)
model <- lm(Men200m ~ Year)

plot(x=Year, y=Men200m,
     xlab="Year", ylab="Men's 200m time (secs)",
     panel.last = lines(sort(Year), fitted(model)[order(Year)]))

```
Display model results.
Display analysis of variance table.
```{r}
summary(model)
```
```{r}
anova(model)
detach(sprinters)
```
# Highway sign reading distance and driver age
Load the signdist data.
Fit a simple linear regression model with y = Distance and x = Age.
Display a scatterplot of the data with the simple linear regression line.
```{r}
signdist <- read.table("Data2/signdist.txt", header=T)
attach(signdist)

model <- lm(Distance ~ Age)

plot(x=Age, y=Distance,
     xlab="Age", ylab="Distance",
     panel.last = lines(sort(Age), fitted(model)[order(Age)]))

```
Display model results.
```{r}
summary(model)
```
Calculate confidence intervals for the slope.
```{r}
confint(model, parm="Age", level=0.95)
```

```{r}
confint(model, parm="Age", level=0.99)

```
```{r}
detach(signdist)
```
# Handspan and height
Load the handheight data.
```{r}
handheight <- read.table("Data2/handheight.txt", header=T)
attach(handheight)
head(handheight, n=10)
```
Fit a simple linear regression model with y = Height and x = Handspan
Display a scatterplot of the data with the simple linear regression line.
```{r}
model <- lm(Height ~ HandSpan)

plot(x=HandSpan, y=Height,
     xlab="HandSpan", ylab="Height",
     panel.last = lines(sort(HandSpan), fitted(model)[order(HandSpan)]))
```
Display model results.
```{r}
summary(model)
```
Display analysis of variance table.
```{r}
anova(model)

detach(handheight)
```
* The residual standard deviation S is 2.744 and this estimates the standard deviation of the errors.
    r2 = (SSTO-SSE) / SSTO = SSR / (SSR+SSE) = 1500.1 / (1500.1+1242.7) = 1500.1 / 2742.8 = 0.547 or 54.7%. * * The interpretation is that handspan differences explain 54.7% of the variation in heights.
* The value of the F statistic is F = 199.2 with 1 and 165 degrees of freedom, and the p-value for this F statistic is 0.000. Thus we reject the null hypothesis H0 : ??1 = 0 because the p-value is so small. In other words, the observed relationship is statistically significant.

# Checking account deposits

Display usual analysis of variance table.
Load the newaccounts data.
```{r}
newaccounts <- read.table("Data2/newaccounts.txt", header=T)
head(newaccounts)
attach(newaccounts)
```
Fit a simple linear regression model with y = New and x = Size
Display a scatterplot of the data with the simple linear regression line.
```{r}
model <- lm(New ~ Size)

plot(x=Size, y=New,
     xlab="Size of minimum deposit", ylab="Number of new accounts",
     panel.last = lines(sort(Size), fitted(model)[order(Size)]))
```
Display model results.
```{r}
summary(model)
```
Display lack of fit analysis of variance table.
```{r message=FALSE, warning=FALSE}
library(alr3) # alr3 package must be installed first
pureErrorAnova(model) # Lack of fit anova table
```
NOTE: The F value for Size uses MSPE in its denominator.
So, F value for Size is 5141.3 / 229.6 = 22.393.
Thus it differs from the F value for Size in the usual anova table: 
```{r}
anova(model)
# NOTE: Here the F value for Size uses MSE in its denominator.
# So, F value for Size is 5141.3 / 1638.0 = 3.1389.
detach(newaccounts)
```

