---
title: "Lesson 14: Time Series & Autocorrelation"
output: html_notebook
---
https://onlinecourses.science.psu.edu/stat501/node/418/
# Google stock (autoregression model)

    Use the read.zoo function in the zoo package to load the google_stock data in time series format.
    Create a time series plot of the data.
    Load the google_stock data in the usual way using read-table.
    Use the ts function to convert the price variable to a time series.
    Create a plot of partial autocorrelations of price.
    Calculate a lag-1 price variable (note that the lag argument for the function is -1, not +1).
    Create a scatterplot of price vs lag1price.
    Use the ts.intersect function to create a dataframe containing price and lag1price.
    Fit a simple linear regression model of price vs lag1price (a first-order autoregression model).

```{r}
library(zoo)
google <- read.csv("DataL14/google_stock.csv", header=T)
google.ts <- ts(google$Price)
head(google.ts)
plot(google.ts)
```
```{r}
pacf(google.ts)
```
```{r}
lag1price <- lag(google.ts, -1)
plot(google.ts ~ lag1price, xy.labels=F)
```
```{r}
lagdata <- ts.intersect(google.ts, lag1price, dframe=T)
summary(lm(google.ts ~ lag1price, data=lagdata))
```
# Earthquakes (autoregression model)

    Load the earthquakes data.
    Create a time series plot of the data.
    Use the ts function to convert the Quakes variable to a time series.
    Create a plot of partial autocorrelations of Quakes.
    Calculate lag-1, lag-2, and lag-3 Quakes variables.
    Use the ts.intersect function to create a dataframe containing Quakes and the three lag variables.
    Fit a multiple linear regression model of Quakes versus the three lag variables (a third-order autoregression model).

```{r}
earthquakes <- read.table("DataL14/earthquakes.txt", header=T)
attach(earthquakes)

plot(Year, Quakes, type="b")
```
```{r}
Quakes.ts <- ts(Quakes)
pacf(Quakes.ts)
```
```{r}
lag1Quakes <- lag(Quakes.ts, -1)
lag2Quakes <- lag(Quakes.ts, -2)
lag3Quakes <- lag(Quakes.ts, -3)

lagdata <- ts.intersect(Quakes.ts, lag1Quakes, lag2Quakes, lag3Quakes, dframe=T)
summary(lm(Quakes.ts ~ lag1Quakes + lag2Quakes + lag3Quakes, data=lagdata))
detach(earthquakes)
```
# Blaisdell company (regression with autoregressive errors)

    Load the blaisdell data.
    Fit a simple linear regression model of comsales vs indsales.
    Use the dwt function in the car package to conduct the Durbin-Watson test on the residuals.
    Conduct the Ljung-Box test on the residuals.
    Perform the Cochrane-Orcutt procedure to transform the variables.
    Forecast comsales for period 21 when indsales are projected to be $175.3 million.
    Perform the Hildreth-Lu procedure to transform the variables.
    Perform the first differences procedure to transform the variables.
```{r}
blaisdell <- read.table("DataL14/Blaisdell_Company.csv", sep=",",  header=T)
attach(blaisdell)
head(blaisdell)

model.1 <- lm(comsales ~ indsales)
summary(model.1)
```
```{r}
# Durbin-Watson Test
library(car)
dwt(model.1)
```
```{r}
# Ljung-Box Q Test
Box.test(residuals(model.1), lag = 1, type = "Ljung")
```
```{r}
# Cochrane-Orcutt Procedure
res.ts <- ts(residuals(model.1))
lag1res <- lag(res.ts, -1)
lagdata1 <- ts.intersect(res.ts, lag1res)
acp <- coef(lm(res.ts ~ lag1res -1, data=lagdata1)) # 0.6311636
y.ts <- ts(comsales)
x.ts <- ts(indsales)
lag1y <- lag(y.ts, -1)
lag1x <- lag(x.ts, -1)
y.co <- y.ts-acp*lag1y
x.co <- x.ts-acp*lag1x
model.2 <- lm(y.co ~ x.co)
summary(model.2)
```
```{r}
dwt(model.2)
```
```{r}
b0 <- coef(model.2)[1]/(1-acp) # -1.068524
sqrt(vcov(model.2)[1,1])/(1-acp) # se = 0.4533986
b1 <- coef(model.2)[2] # 0.1737583

```
```{r}
fit.20 <- b0+b1*indsales[20] # 28.76577
fit.20
res.20 <- comsales[20]-fit.20 # 0.01422919
fit.21 <- b0+b1*175.3 # 29.3913
forecast.21 <- fit.21+acp*res.20 # 29.40028
fit.21
```
```{r}
# Hildreth-Lu Procedure
sse <- vector()
for(i in 1:90){
  y.hl = y.ts-(0.09+0.01*i)*lag1y
  x.hl = x.ts-(0.09+0.01*i)*lag1x
  sse[i] <- sum(residuals(lm(y.hl ~ x.hl))^2)
}
acp <- 0.09+0.01*which.min(sse) # 0.96
y.hl = y.ts-acp*lag1y
x.hl = x.ts-acp*lag1x
model.3 <- lm(y.hl ~ x.hl)
summary(model.3)
```

